At first you'll need to:

1) Download the "report.pg_matrix_test.tsv", "experiment_annotations_test.tsv" and "all_proteins.csv" in your working directory

2) install all the necessary packages for our work

```{r eval=FALSE}
install.packages(c("mpwR", "renv", "tidyverse", "reshape2"))
renv::install("bioc::ComplexHeatmap")
renv::install("bioc::limma")
renv::install("bioc::MSnbase")
renv::install("bioc::SummarizedExperiment")
renv::install("bioc::cmapR")
renv::install("bioc::ConsensusClusterPlus")

renv::install("Nesvilab/FragPipeAnalystR")

renv::install("nicolerg/ssGSEA2")

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("AnnotationDbi")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("GO.db")
```

Congratulations! Now you are ready to go through the tutorial!

Load your libraries

```{r eval=FALSE}
library(FragPipeAnalystR)
library(tidyverse)
library(UpSetR)
library(reshape2)
library(org.Hs.eg.db)
library(AnnotationDbi)
library(GO.db)
```

We will start from creating the SummarizedExperiment object which contains the information about abundances of different proteins in different samples and the ex^perimental setup.

```{r eval=FALSE}
se <- make_se_from_files(
  "report.pg_matrix_test.tsv",
  "experiment_annotations_test.tsv",
  type = "DIA"
)
```

Now let's perform some quality control (QC) steps.

```{r eval=FALSE}
pdf("PCA_test.pdf", width = 14, height = 8); plot_pca(se); dev.off()
pdf("Correlation_test.pdf", width = 10, height = 10); plot_correlation_heatmap(se); dev.off()
pdf("Missing_test.pdf", width = 10, height = 10); plot_missval_heatmap(se); dev.off()
pdf("Features_test.pdf", width = 20, height = 10); plot_feature_numbers(se); dev.off()
```

What do you think about the quality of this data?

Let's explore the missing values a little bit more

```{r eval=FALSE}
df<-assay(se)
MVs <- rowSums(is.na(df)) / ncol(df)
MVs<-MVs[which(MVs!=0)]
MVs<-MVs[which(MVs!=1)]

intensities <- rowMeans(df, na.rm = TRUE)
intensities<-intensities[which(names(intensities) %in% names(MVs))]


data<-data.frame("Intensity"=intensities, "MVs"=MVs)
ggplot(data, aes(x=Intensity, y=MVs)) + 
  geom_point()
```


What does this plot tell you about your missing values?

Let's finally impute them!

```{r eval=FALSE}
se_imp<-manual_impute(se)

df_imp<-assay(se_imp)

# Extract condition names
conds <- sub("_[^_]+$", "", colnames(df))

# Split column indices by condition
cond_names <- split(colnames(df), conds)

# For each condition, count NAs per row
na_counts <- sapply(cond_names, function(cols) {
  rowSums(is.na(df[, cols, drop = FALSE]))
})

# Convert to data frame
na_counts <- as.data.frame(na_counts)

cond_sizes <- sapply(cond_names, length)

all_na  <- sweep(na_counts, 2, cond_sizes, `==`)
two_na <- sweep(na_counts, 2, cond_sizes - 1, `==`)

rows_with_allna <- apply(all_na, 1, any)
rows_with_twona <- apply(two_na, 1, any)

# Row indices in original df
rows_to_fix <- which(rowSums(all_na) > 0)
# Loop over rows
for (r in rows_to_fix) {
  # For each condition that is all-NA in this row
  conds_all_na <- colnames(all_na)[all_na[r, ]]
  
  for (cond in conds_all_na) {
    cols <- cond_names[[cond]]
    df_imp[r, cols] <- 0   # replace NA with 0
  }
}


rows_to_fix <- which(rowSums(two_na) > 0)
# Loop over rows
for (r in rows_to_fix) {
  conds_two_na <- colnames(two_na)[two_na[r, ]]
  
  for (cond in conds_two_na) {
    cols <- cond_names[[cond]]
    
    # Replace ONLY the NA values
    na_idx <- is.na(df[r, cols])
    df_imp[r, cols][na_idx] <- 0
  }
}

assay(se_imp)<-df_imp

pdf("PCA_after_imputation_test.pdf", width = 14, height = 8); plot_pca(se_imp); dev.off()

```

We need to check if our imputation did not mess up with our data (did not skew significantly our intensity distributions or something)

```{r eval=FALSE}
obs <- as.vector(as.matrix(df))
obs <- obs[!is.na(obs)]
imp <- as.vector(as.matrix(df_imp))
imp <- imp[is.na(as.matrix(df))]
df_plot <- data.frame(
  Intensity = c(obs, imp),
  Type = c(rep("Observed", length(obs)), rep("Imputed", length(imp)))
)

# Plot densities
ggplot(df_plot, aes(x = Intensity, color = Type, fill = Type)) +
  geom_density(alpha = 0.3) +
  scale_color_manual(values = c("blue", "red")) +
  scale_fill_manual(values = c("blue", "red")) +
  xlab("log2 intensity") + ylab("Density") +
  theme_minimal()
```

So what do you think? Looks good or not? Why?

```{r eval=FALSE}
frac_missing <- rowSums(is.na(df)) / ncol(df)
mean_intensity <- rowMeans(df_imp)

plot(mean_intensity, frac_missing,
     xlab = "Mean log2 intensity (after imputation)",
     ylab = "Fraction originally missing",
     pch = 16, col = rgb(0,0,1,0.5))
```

In order to compare protein abundances between different samples we need to normalize their medians. But how does it look before normalization?

```{r eval=FALSE}
# Flatten assay data for plotting
mat <- assay(se_imp)  # your SummarizedExperiment after imputation
df_plot <- melt(mat)
colnames(df_plot) <- c("Protein", "Sample", "Intensity")
df_plot$Condition <- sub("_(?!.*_).*", "", df_plot$Sample, perl = TRUE)

ggplot(df_plot, aes(x = Sample, y = Intensity, fill = Condition)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("log2 intensity") +
  xlab("Sample")
```

Now let's normalize and check how it looks after.

```{r eval=FALSE}
se_imp_norm<-MD_normalization(se_imp)

mat <- assay(se_imp_norm)  # your SummarizedExperiment after imputation
df_plot <- melt(mat)
colnames(df_plot) <- c("Protein", "Sample", "Intensity")
df_plot$Condition <- sub("_(?!.*_).*", "", df_plot$Sample, perl = TRUE)

ggplot(df_plot, aes(x = Sample, y = Intensity, fill = Condition)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("log2 intensity") +
  xlab("Sample")
```

Finally, my friends, the moment of truth! We are going to perform differential abundance analysis!!

```{r eval=FALSE}
final_contrasts <- c("C13.A_vs_C13.B", "C13.A_vs_C13.C", "C13.B_vs_C13.C")

de_results <- test_limma(se_imp_norm, type = "manual", test=final_contrasts) 
de_results_updated <- add_rejections(de_results)

for (ct in final_contrasts) {
  pdf(file.path(paste0(ct, ".pdf")))
  print(plot_volcano(de_results_updated, ct))
  dev.off()
}

for (ct in final_contrasts) {
  
  padj_col <- paste0(ct, "_p.adj")
  
  sig <- de_results_updated@elementMetadata[
    !is.na(de_results_updated@elementMetadata[[padj_col]]) &
      de_results_updated@elementMetadata[[padj_col]] < 0.05,
  ]
  
  write.csv(
    sig[,c(colnames(sig)[1:8],colnames(sig)[grepl(ct, colnames(sig))])],
    file = file.path(paste0(ct, ".csv")),
    row.names = FALSE
  )
}


```


I'll show you another fun type of plot:

```{r eval=FALSE}

upset_prots<-c()
upset_conds<-c()

for (ct in final_contrasts) {
  
  padj_col <- paste0(ct, "_p.adj")
  sig <- de_results_updated@elementMetadata[
    !is.na(de_results_updated@elementMetadata[[padj_col]]) &
      de_results_updated@elementMetadata[[padj_col]] < 0.05 &
      de_results_updated@elementMetadata[[paste0(ct, "_significant")]]==TRUE,
  ]$Protein.Group
  upset_prots <- c(upset_prots,sig)
  upset_conds<-c(upset_conds, rep(ct, length(sig)))
} 
upset_df<-data.frame("Proteins" = upset_prots, "Contrasts" = upset_conds)

upset_list<-list("C13.A_vs_C13.B" = upset_df$Proteins[which(upset_df$Contrasts == unique(upset_df$Contrasts)[1])], "C13.A_vs_C13.C" = upset_df$Proteins[which(upset_df$Contrasts == unique(upset_df$Contrasts)[2])], "C13.B_vs_C13.C" = upset_df$Proteins[which(upset_df$Contrasts == unique(upset_df$Contrasts)[3])])

plot_Upset(upset_list)

```

Now after we saw which proteins are differentially abundant between conditions let's see which GO terms are enriched between conditions:

```{r eval=FALSE}
#Helper functions
first_uniprot <- function(x) {
  
  x <- as.character(x)
  
  sapply(
    strsplit(x, ";"),
    `[`,
    1
  )
}

run_go_binomial <- function(genes,
                            go_annot,
                            background) {
  
  # Keep first protein accession only
  genes <- first_uniprot(genes)
  background <- first_uniprot(background)
  
  # Remove duplicates
  genes <- unique(genes)
  background <- unique(background)
  
  # Number of proteins
  n <- length(genes)
  
  # GO mappings
  bg_go <- go_annot[
    go_annot$UNIPROT %in% background,
  ]
  
  fg_go <- go_annot[
    go_annot$UNIPROT %in% genes,
  ]
  
  # GO term counts
  bg_counts <- table(bg_go$GO)
  fg_counts <- table(fg_go$GO)
  
  # Results table
  res <- data.frame(
    GO = names(fg_counts),
    k = as.integer(fg_counts)
  )
  
  # Background frequency
  res$bg_freq <- as.numeric(
    bg_counts[res$GO]
  ) / length(background)
  
  # Binomial enrichment test
  res$pval <- pbinom(
    res$k - 1,
    n,
    res$bg_freq,
    lower.tail = FALSE
  )
  
  # Multiple testing correction
  res$adj.p <- p.adjust(
    res$pval,
    method = "BH"
  )
  
  # Sort results
  res <- res[
    order(res$adj.p),
  ]
  
  rownames(res) <- NULL
  
  res
}

annotate_go_terms <- function(go_res) {
  
  term_names <- AnnotationDbi::select(
    GO.db,
    keys = go_res$GO,
    columns = c(
      "TERM",
      "ONTOLOGY"
    ),
    keytype = "GOID"
  )
  
  go_res <- merge(
    go_res,
    term_names,
    by.x = "GO",
    by.y = "GOID",
    all.x = TRUE
  )
  
  # Human-readable ontology names
  go_res$ONTOLOGY <- factor(
    go_res$ONTOLOGY,
    levels = c(
      "MF",
      "BP",
      "CC"
    ),
    labels = c(
      "Molecular Function",
      "Biological Process",
      "Cellular Component"
    )
  )
  
  go_res
}

plot_go <- function(go_res,
                    top_n = 20,
                    alpha = 0.05,
                    title = "") {
  
  # Significant terms only
  sig <- go_res[
    go_res$adj.p < alpha,
  ]
  
  if (nrow(sig) == 0) {
    return(NULL)
  }
  
  # Sort by adjusted p-value
  sig <- sig[
    order(sig$adj.p),
  ]
  
  # Keep top terms
  sig <- sig[
    1:min(top_n, nrow(sig)),
  ]
  
  # Avoid log10(0)
  sig$adj.p[
    sig$adj.p == 0
  ] <- .Machine$double.xmin
  
  # Labels
  go_labels <- paste0(
    sig$TERM,
    " (",
    sig$GO,
    ")"
  )
  
  # Dynamic left margin
  max_chars <- max(
    nchar(go_labels)
  )
  
  left_mar <- min(
    60,
    max(
      20,
      ceiling(max_chars * 0.6)
    )
  )
  
  par(
    mar = c(
      5,
      left_mar,
      4,
      2
    )
  )
  
  # Plot
  barplot(
    -log10(sig$adj.p),
    names.arg = go_labels,
    horiz = TRUE,
    las = 2,
    main = title,
    xlab = "-log10(adj.p)",
    col = "steelblue"
  )
}

save_go_category <- function(go_res,
                             ontology_name,
                             contrast,
                             direction) {
  
  # Filter ontology
  sub <- go_res[
    go_res$ONTOLOGY == ontology_name,
  ]
  
  # Skip empty results
  if (nrow(sub) == 0) {
    return(NULL)
  }
  
  # Ontology abbreviation
  ont_short <- c(
    "Biological Process" = "BP",
    "Molecular Function" = "MF",
    "Cellular Component" = "CC"
  )[ontology_name]
  
  # -----------------------------------------------------
  # Save plot
  # -----------------------------------------------------
  
  pdf(
    paste0(
      contrast,
      "_",
      direction,
      "_",
      ont_short,
      ".pdf"
    ),
    width = 14,
    height = 8
  )
  
  plot_go(
    sub,
    title = paste(
      contrast,
      direction,
      ontology_name
    )
  )
  
  dev.off()
  
  # -----------------------------------------------------
  # Save table
  # -----------------------------------------------------
  
  write.csv(
    sub,
    paste0(
      contrast,
      "_",
      direction,
      "_",
      ont_short,
      ".csv"
    ),
    row.names = FALSE
  )
}

# Building GO annotation table

go_table <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = keys(
    org.Hs.eg.db,
    keytype = "UNIPROT"
  ),
  columns = c(
    "GO",
    "ONTOLOGY"
  ),
  keytype = "UNIPROT"
)

# Remove missing GO annotations
go_table <- go_table[
  !is.na(go_table$GO),
]


# ---------------------------------------------------------
# Background universe
# ---------------------------------------------------------


all_proteins <- read.csv(
  "all_proteins.csv"
)

background <- first_uniprot(
  all_proteins$Protein.Ids
)

background <- unique(background)


# ---------------------------------------------------------
# GO enrichment loop
# ---------------------------------------------------------

ontologies <- c(
  "Biological Process",
  "Molecular Function",
  "Cellular Component"
)

for (contrast in final_contrasts) {
  
  cat(
    "\n==============================\n"
  )
  
  cat(
    "Running contrast:",
    contrast,
    "\n"
  )
  
  cat(
    "==============================\n"
  )
  
  # -----------------------------------------------------
  # Read differential expression results
  # -----------------------------------------------------
  
  de <- read.csv(
    paste0(
      contrast,
      ".csv"
    )
  )
  
  # Fold-change column
  fc_col <- paste0(
    contrast,
    "_diff"
  )
  
  # -----------------------------------------------------
  # Upregulated proteins
  # -----------------------------------------------------
  
  up <- de$Protein.Ids[
    de[[fc_col]] > 0
  ]
  
  # -----------------------------------------------------
  # Downregulated proteins
  # -----------------------------------------------------
  
  down <- de$Protein.Ids[
    de[[fc_col]] < 0
  ]
  
  cat(
    "Upregulated proteins:",
    length(up),
    "\n"
  )
  
  cat(
    "Downregulated proteins:",
    length(down),
    "\n"
  )
  
  # -----------------------------------------------------
  # Run enrichment
  # -----------------------------------------------------
  
  go_up <- run_go_binomial(
    up,
    go_table,
    background
  )
  
  go_down <- run_go_binomial(
    down,
    go_table,
    background
  )
  
  # -----------------------------------------------------
  # Annotate GO terms
  # -----------------------------------------------------
  
  go_up <- annotate_go_terms(
    go_up
  )
  
  go_down <- annotate_go_terms(
    go_down
  )
  
  # -----------------------------------------------------
  # Save ontology-specific results
  # -----------------------------------------------------
  
  for (ont in ontologies) {
    
    save_go_category(
      go_up,
      ont,
      contrast,
      "UP"
    )
    
    save_go_category(
      go_down,
      ont,
      contrast,
      "DOWN"
    )
  }
  
  cat(
    "Finished:",
    contrast,
    "\n"
  )
}
```

And, yeah... We are done for today, thank you for being here!

